---
title: 'P&S-2025: Lab assignment 4'
author: "Daryna Shevchuk, Mykyta Yagoda, Teodor Talan"
output:
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

## Work breakdown:

-   Teodor Talan: Problem **1**, Problem **2**\
-   Daryna Shevchuk: Problem **3**\
-   Mykyta Yagoda: Problem **4**\

The data for Problems 1--3 are generated as follows: set
$$
a_k := \{ k \ln (k^{2n} + \pi) \}, \quad k \ge 1,
$$
where
$$
\{x\} := x - \lfloor x \rfloor
$$
is the fractional part of a number $x$, and $n$ is your id number.

Sample realizations $X_1, \dots, X_{100}$ and $Y_1, \dots, Y_{50}$ from the hypothetical normal distributions
$N(\mu_1, \sigma_1^2)$ and $N(\mu_2, \sigma_2^2)$, respectively, are obtained as
$$
x_k = \Phi^{-1}(a_k), \quad k = 1, \dots, 100,
$$
$$
y_l = \Phi^{-1}(a_{l+100}), \quad l = 1, \dots, 50,
$$
where $\Phi$ is the cumulative distribution function of $N(0,1)$, and $\Phi^{-1}$ is its inverse.

\textbf{Instructions.} In Problems 1--3, test $H_0$ versus $H_1$. To this end:

- point out which standard statistical test you use and justify your choice;
- indicate the general form of the rejection region for the test of $H_0$ versus $H_1$ at the significance level $\alpha = 0.05$;
- determine whether $H_0$ should be rejected at the significance level $0.05$;
- indicate the $p$-value of the test and explain whether you would reject $H_0$ for this value of $p$ and why.


### Problem 1: 

$$
H_0 : \mu_1 = \mu_2 \quad \text{vs.} \quad H_1 : \mu_1 \neq \mu_2; \quad \sigma_1^2 = \sigma_2^2 = 1.
$$

### Problem 2:

$$
H_0 : \sigma_1^2 = \sigma_2^2 \quad \text{vs.} \quad H_1 : \sigma_1^2 > \sigma_2^2; \quad \mu_1 \text{ and } \mu_2 \text{ are unknown.}
$$

### Problem 3:

Using Kolmogorov--Smirnov test, check if

(a) $\{x_k\}_{k=1}^{100}$ are normally distributed (with parameters calculated from the sample);

```{r}
n <- 23

a_k <- function(k) {
  (k*log(k^2*n+pi)) %% 1
}

ak.data <- sapply(1:150, a_k)
x <- qnorm(ak.data[1:100])
y <- qnorm(ak.data[101:150])

mu_hat <- mean(x)
sd_hat <- sd(x)

mu_hat
sd_hat
```

```{r}
ks_res <- ks.test(x, "pnorm", mean = mu_hat, sd = sd_hat)
ks_res
```
We consider the hypothesis
\[
H_0: \{x_k\}_{k=1}^{100} \text{ are observations from } \mathcal{N}(\mu,\sigma^2)
\]
with $(\mu,\sigma^2)$ estimated from the sample, against the general alternative
\[
H_1: \{x_k\}_{k=1}^{100} \text{ do not follow } \mathcal{N}(\mu,\sigma^2).
\]

\textbf{Test and justification.}
To test $H_0$ vs.\ $H_1$ we use the one-sample Kolmogorov--Smirnov goodness-of-fit test.  
This is a standard nonparametric test that compares the empirical distribution function $F_n(x)$ 
of the sample with the theoretical distribution function $F(x;\hat\mu,\hat\sigma)$ of the fitted normal law.
It is appropriate here because we want to check whether the whole distribution of the data agrees with a normal model,
not just particular moments such as the mean or variance.

\textbf{General form of the rejection region.}
Let
\[
D_n = \sup_x \bigl| F_n(x) - F(x;\hat\mu,\hat\sigma) \bigr|
\]
be the Kolmogorov--Smirnov test statistic.  
For a test of level $\alpha = 0.05$ the rejection region has the general form
\[
R_\alpha = \{ D_n > c_{0.05} \},
\]
where $c_{0.05}$ is the critical value (or, equivalently, we reject $H_0$ when the p-value is smaller than $\alpha$).

\textbf{Numerical results and decision at } $\alpha = 0.05$.
For our sample we obtain the parameter estimates
\[
\hat{\mu} = -0.001800432, \quad \hat{\sigma} = 1.034129,
\]
and the Kolmogorov--Smirnov statistic
\[
D = 0.045628, \quad p\text{-value} = 0.9853.
\]
Since $p\text{-value} = 0.9853 > \alpha = 0.05$, we do not reject $H_0$ at the 5\% significance level.

\textbf{Conclusion in terms of the p-value.}
Because the p-value is very large (0.9853), the observed discrepancy between the empirical distribution function 
and the fitted normal distribution is well within what could reasonably be expected under $H_0$.  
Therefore, we do not reject the null hypothesis, and we conclude that the sample 
$\{x_k\}_{k=1}^{100}$ is consistent with a normal distribution with parameters estimated from the data.


(b) $\{|x_k|\}_{k=1}^{100}$ are exponentially distributed with $\lambda = 1$;

```{r}
abs_x <- abs(x)

ks_res <- ks.test(abs_x, "pexp", rate = 1)
ks_res
```
We consider the hypothesis
\[
H_0: \{|x_k|\}_{k=1}^{100} \text{ are observations from the exponential distribution } \operatorname{Exp}(1),
\]
against the general alternative
\[
H_1: \{|x_k|\}_{k=1}^{100} \text{ do not follow } \operatorname{Exp}(1).
\]

\textbf{Test and justification.}
To test $H_0$ vs.\ $H_1$ we use the one-sample Kolmogorov--Smirnov goodness-of-fit test.  
This is a standard nonparametric test that compares the empirical distribution function $F_n(x)$ 
of the sample $\{|x_k|\}$ with the theoretical distribution function $F(x)$ of the exponential law with rate $\lambda = 1$:
\[
F(x) = 1 - e^{-x}, \quad x \ge 0.
\]
The test is appropriate here because we want to check whether the entire distribution of the data agrees with the exponential model, 
not just particular characteristics such as the mean.

\textbf{General form of the rejection region.}
Let
\[
D_n = \sup_x \bigl| F_n(x) - F(x) \bigr|
\]
be the Kolmogorov--Smirnov test statistic.  
For a test of level $\alpha = 0.05$ the rejection region has the general form
\[
R_\alpha = \{ D_n > c_{0.05} \},
\]
where $c_{0.05}$ is the critical value corresponding to the Kolmogorov--Smirnov distribution (or, equivalently, we reject $H_0$ when the p-value is smaller than $\alpha$).

\textbf{Numerical results and decision at } $\alpha = 0.05$.
For our sample $\{|x_k|\}_{k=1}^{100}$ the Kolmogorov--Smirnov test yields the statistic
\[
D = 0.10893, \quad p\text{-value} = 0.1862.
\]
Since $p\text{-value} = 0.1862 > \alpha = 0.05$, we do not reject $H_0$ at the 5\% significance level.

\textbf{Conclusion in terms of the p-value.}
Because the p-value is relatively large ($0.1862$), the observed discrepancy between the empirical distribution function 
and the exponential distribution with rate $\lambda = 1$ is within the range of what could be expected under $H_0$.  
Therefore, we do not reject the null hypothesis, and conclude that the sample 
$\{|x_k|\}_{k=1}^{100}$ is consistent with an $\operatorname{Exp}(1)$ distribution.

(c) $\{x_k\}_{k=1}^{100}$ and $\{y_l\}_{l=1}^{50}$ have the same distributions.

```{r}
ks_x_y <- ks.test(x, y)
ks_x_y
```
We consider the hypotheses
\[
H_0: \{x_k\}_{k=1}^{100} \text{ and } \{y_l\}_{l=1}^{50} \text{ are drawn from the same distribution},
\]
against
\[
H_1: \{x_k\}_{k=1}^{100} \text{ and } \{y_l\}_{l=1}^{50} \text{ do not have the same distribution}.
\]

\textbf{Test and justification.}
To test $H_0$ versus $H_1$, we use the two-sample Kolmogorov--Smirnov test.
This is a standard nonparametric procedure for comparing two independent samples 
in order to determine whether they come from the same underlying distribution.
The test is sensitive to differences in both the location and the shape of distributions.

\textbf{General form of the rejection region.}
Let
\[
D_{n,m} = \sup_x \bigl| F_n(x) - G_m(x) \bigr|
\]
be the Kolmogorov--Smirnov statistic, where $F_n$ and $G_m$ 
are the empirical distribution functions of the samples $\{x_k\}$ and $\{y_l\}$.
For a test of level $\alpha = 0.05$, the rejection region has the general form
\[
R_\alpha = \{ D_{n,m} > c_{0.05} \},
\]
where $c_{0.05}$ is the critical value corresponding to the two-sample Kolmogorov--Smirnov distribution
(or, equivalently, $H_0$ is rejected when the p-value is smaller than $\alpha$).

\textbf{Numerical results and decision at } $\alpha = 0.05$.
For the two-sample Kolmogorov--Smirnov test, we obtain
\[
D = 0.16, \quad p\text{-value} = 0.3503.
\]
Since $p\text{-value} = 0.3503 > \alpha = 0.05$, we do not reject $H_0$ at the 5\% significance level.

\textbf{Conclusion in terms of the p-value.}
Because the p-value is not small ($0.3503$), the observed difference between the empirical distribution functions 
of $\{x_k\}$ and $\{y_l\}$ can be explained by random variation under $H_0$.  
Therefore, we do not reject the null hypothesis and conclude that the two samples
$\{x_k\}_{k=1}^{100}$ and $\{y_l\}_{l=1}^{50}$ are consistent with being drawn from the same distribution.

### Problem 4:

In this task you’ll practice fitting the regression line to some real-life features and analyzing the results.
The file \texttt{data.csv} contains data on students, specifically their study time and corresponding marks.
Your tasks are as follows:

(a) Create a scatter plot of Marks vs. Study Time and provide brief comments;

(b) Fit a linear regression model using Marks as the dependent variable and Study Time as the independent variable.
Explain shortly the process of deriving the regression equation;

(c) Evaluate the goodness-of-fit for the fitted line;

(d) Suggest a way to test whether the Study Time is significant in predicting Marks. Find the corresponding test
statistic, specify its distribution. Find the $p$-value of the test and make a conclusion;

(e) If Alice studies for approximately $8$ hours, what grade can we predict for her?

(f) Suggest up to three ideas to potentially improve prediction accuracy.


